Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-04-26 drafts/hybrid_zone.tex
minor fixen

The process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
% Gene flow between formerly isolated populations is possible if reproductive isolation is not complete.
However, when formerly isolated populations come into contact before reproductive isolation is complete, some gene flow is possible.
%As two such populations interbreed , each portion of each individual's genome 
%can be labeled according to which ancestral population is is inherited from, and local 
Interbreeding and migration between such populations  
% YB It seems like the fact that such ancestry can be labeled is more important later, but probably shouldn't be a focus here. 
creates a gradient of alleles derived from the two populations across geographic space, %YB removed `assignments' as it seems like it is the ancestry, not its assignment, that is of interest
centered on their point of contact \citep[reviewed by][]{Barton1985}. 
If the populations are sufficiently diverged, 
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal {junctions} in ancestry \citep{Fisher1954, Chapman2002, baird2003distribution}. 
% Limited opportunities for recombination in recent hybrid zones generates substantial correlations in ancestry (and hence, genotype) between distant markers, and
These patterns of correlated ancestry in admixed populations have been previously used to infer histories of hybridization and admixture \citep[e.g.][]{Gravel2012,Hellenthal2014,Harris2014,sedghifar2015spatial}.

%The mixing of genomes across space, combined with selection, results in the formation of genetic clines. 
%These patterns have been the traditional focus of both empirical and theoretical studies into hybrid zones. 
%While here we focus on hybrid zones maintained by selection against hybrids 
-%- \emph{tension zones}, as reviewed in \citet{Barton1985} --  % removed \citet{Bazykin1969}, it's cited later
%clines may also be due to differential selection across environmental gradients \citep{Haldane1948,fisher1950frequencies,hanson1966effects},
%% generating \emph{ecotones}.  % I think ecotone means something somewhat different: see Barton & Hewitt 1985, p. 132, for instance
%Although these are in principle different,  \linelabel{ll:all_clines_are_the_same}
%theoretical and simulation studies show that
%genomic clines seen in the two situations are effectively indistinguishable without additional information \citep{may1975gene,Barton1993,Kruuk1999},
%with their shape determined mostly by the mean fitness of the heterozygote \citep{Slatkin1973}.
% consistently show ecotones and tension zones have very similar properties \citep{Slatkin1973, barton1986barrier,Kruuk1999}, 
% \plr{(I can't find discussion of similarity of different types of clines in \citet{barton1986barrier}: put the citation back in if I missed it.)}
%% don't think we need this next bit
% The primary difference being that the center of a tension zone fluctuates over time \citep{barton1979dynamics} 
% because stochastic changes in allele frequency can shit the cline center, 
% while centers of ecotones do not change unless the environmental gradient changes.

@@

Using a combination of theory and simulated hybrid zones, we describe genome-wide patterns of coancestry 
as they relate to hybrid zone age, genetic distance from selected locus and geographic distance from hybrid zone center. 
\alisa{We extend the framework presented by \cite{Sedghifar2015} to describe how lineages move across a zone of secondary contact. We consider the influence of selection against heterozygotes at a single locus on the movement of lineages across space, conditioning on the frequencies of the selected allele and allowing lineages to recombine between selected background } (analogous to the structured coalescent applied to selection, as in \citet{hudson1988coalescent}).
%In \citet{sedghifar2015spatial} we described how lineages move across a neutral zone of secondary contact,
%obtaining predictions for distributions of lengths of blocks of ancestry, which we fit to several datasets.
%In this work, we extend the same framework to the case of selection against heterozygotes at a single locus,
%by describing how lineages move conditional on the frequencies of the selected allele
%and allowing lineages to recombine between selected background 
%(analogous to the structured coalescent applied to selection, as in \citet{hudson1988coalescent}).

@@

We can describe this process formally in It\^o notation:
with $B_\tau$ a standard Brownian motion,
$T_B(\tau)$ the most recent time before $\tau$ that $Z_\tau = B$, and likewise for $T_A$,
\begin{align}
    \begin{aligned} \label{eqn:lineage_motion}
        dX_\tau &= \sigma dB_\tau + \begin{cases}
            \sigma^2 \frac{d}{dx} \log(p(X_\tau,\tau)) d\tau \qquad & \text{if } Z_\tau = A \\
             \sigma^2 \frac{d}{dx} \log(1-p(X_\tau,\tau)) d\tau \qquad & \text{if } Z_\tau = B 
        \end{cases} \\
        % \P\{ Z_{\tau+\epsilon} = B \given Z_\tau = A \} &= \epsilon r (1-p(X_\tau,\tau)) + O(\epsilon^2) \\
        % \P\{ Z_{\tau+\epsilon} = A \given Z_\tau = B \} &= \epsilon r p(X_\tau,\tau) + O(\epsilon^2)  .
        \P\{ T_B(\tau) &> \tau+u \given Z_\tau = A \} = \exp\left( - r \int_\tau^{\tau+u} (1-p(X_s,s)) ds \right) \\
        \P\{ T_A(\tau) &> \tau+u \given Z_\tau = B \} = \exp\left( - r \int_\tau^{\tau+u} p(X_s,s) ds \right) .
    \end{aligned}
\end{align}
In the first expression, giving the distribution of how the lineage location changes,
the first term ($\sigma dB_\tau$) is Brownian noise driven by dispersal,
and the second is the mean displacement, which moves the lineage ``downhill'' towards its selected allele's ancestral range on either $-\log p$ or $-\log (1-p)$,
depending on which selected allele the lineage is linked to.
(In two dimensions, $d/dx$ is replaced by the gradient.)
\plr{
The second expression gives the probability that a lineage on the $A$ background at time $\tau$
does not switch to the $B$ background across $u$ generations,
i.e., the probability that there was no recombination between the lineage and the selected site in those $u$ generations.
Since the rate of recombination between the two is $r$,
and the probability the lineage is in a heterozygote at time $s$ 
is equal to the local proportion of $B$ alleles, $1-p(X_s,s)$,
the mean number of recombinations in heterozygotes across that time period is $r \int_\tau^{\tau+u} (1-p(X_s,s)) ds$;
assuming recombinations are Poisson, the probability of no such recombinations is the expression given.
The third expression is the same, except for switching $B$ to $A$.
}
% gives the rate at which a lineage switches from an $A$ to $B$ background as a result of recombination,
% which occurs at rate $r$ multiplied by the local proportion of $B$ alleles, which is $(1-p)$.
% The third expression is the corresponding rate for switching from $B$ to $A$.

>>>>

The process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
% Gene flow between formerly isolated populations is possible if reproductive isolation is not complete.
However, when formerly isolated populations come into contact before reproductive isolation is complete, some gene flow is possible.
%As two such populations interbreed , each portion of each individual's genome 
%can be labeled according to which ancestral population is is inherited from, and local 
Interbreeding and migration between such populations  
% YB It seems like the fact that such ancestry can be labeled is more important later, but probably shouldn't be a focus here. 
creates a gradient of alleles derived from the two populations across geographic space, %YB removed `assignments' as it seems like it is the ancestry, not its assignment, that is of interest
centered on their point of contact \citep[reviewed by][]{Barton1985}. 
If the populations are sufficiently diverged, 
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal {junctions} in ancestry \citep{Fisher1954, Chapman2002, baird2003distribution}. 
% Limited opportunities for recombination in recent hybrid zones generates substantial correlations in ancestry (and hence, genotype) between distant markers, and
These patterns of correlated ancestry in admixed populations have been previously used to infer histories of hybridization and admixture \citep[e.g.][]{Gravel2012,Hellenthal2014,Harris2013,sedghifar2015spatial}.

%The mixing of genomes across space, combined with selection, results in the formation of genetic clines. 
%These patterns have been the traditional focus of both empirical and theoretical studies into hybrid zones. 
%While here we focus on hybrid zones maintained by selection against hybrids 
%-- \emph{tension zones}, as reviewed in \citet{Barton1985} --  % removed \citet{Bazykin1969}, it's cited later
%clines may also be due to differential selection across environmental gradients \citep{Haldane1948,fisher1950frequencies,hanson1966effects},
%% generating \emph{ecotones}.  % I think ecotone means something somewhat different: see Barton & Hewitt 1985, p. 132, for instance
%Although these are in principle different,  \linelabel{ll:all_clines_are_the_same}
%theoretical and simulation studies show that
%genomic clines seen in the two situations are effectively indistinguishable without additional information \citep{may1975gene,Barton1993,Kruuk1999},
%with their shape determined mostly by the mean fitness of the heterozygote \citep{Slatkin1973}.
% consistently show ecotones and tension zones have very similar properties \citep{Slatkin1973, barton1986barrier,Kruuk1999}, 
% \plr{(I can't find discussion of similarity of different types of clines in \citet{barton1986barrier}: put the citation back in if I missed it.)}
%% don't think we need this next bit
% The primary difference being that the center of a tension zone fluctuates over time \citep{barton1979dynamics} 
% because stochastic changes in allele frequency can shit the cline center, 
% while centers of ecotones do not change unless the environmental gradient changes.

@@

Using a combination of theory and simulated hybrid zones, we describe genome-wide patterns of coancestry 
as they relate to hybrid zone age, genetic distance from selected locus and geographic distance from hybrid zone center. 
\alisa{We extend the framework presented by \cite{sedghifar2015spatial} to describe how lineages move across a zone of secondary contact. We consider the influence of selection against heterozygotes at a single locus on the movement of lineages across space, conditioning on the frequencies of the selected allele and allowing lineages to recombine between selected background } (analogous to the structured coalescent applied to selection, as in \citet{hudson1988coalescent}).
%In \citet{sedghifar2015spatial} we described how lineages move across a neutral zone of secondary contact,
%obtaining predictions for distributions of lengths of blocks of ancestry, which we fit to several datasets.
%In this work, we extend the same framework to the case of selection against heterozygotes at a single locus,
%by describing how lineages move conditional on the frequencies of the selected allele
%and allowing lineages to recombine between selected background 
%(analogous to the structured coalescent applied to selection, as in \citet{hudson1988coalescent}).

@@

We can describe this Markov process formally in It\^o notation:
with $B_\tau$ a standard Brownian motion,
$T_B(\tau)$ the most recent time before $\tau$ that $Z_\tau = B$, and likewise for $T_A$,
\begin{align}
    \begin{aligned} \label{eqn:lineage_motion}
        dX_\tau &= \sigma dB_\tau + \begin{cases}
            \sigma^2 \frac{d}{dx} \log(p(X_\tau,\tau)) d\tau \qquad & \text{if } Z_\tau = A \\
             \sigma^2 \frac{d}{dx} \log(1-p(X_\tau,\tau)) d\tau \qquad & \text{if } Z_\tau = B 
        \end{cases} \\
        \P\{ Z_{\tau+\epsilon} = B \given Z_\tau = A, \; X_\tau=x \} &= \epsilon r (1-p(x,\tau)) + O(\epsilon^2) \\
        \P\{ Z_{\tau+\epsilon} = A \given Z_\tau = B, \; X_\tau=x \} &= \epsilon r p(x,\tau) + O(\epsilon^2)  .
        % \P\{ T_B(\tau) &> \tau+u \given Z_\tau = A \} = \exp\left( - r \int_\tau^{\tau+u} (1-p(X_s,s)) ds \right) \\
        % \P\{ T_A(\tau) &> \tau+u \given Z_\tau = B \} = \exp\left( - r \int_\tau^{\tau+u} p(X_s,s) ds \right) .
        % \E[ T_B(\tau) &> \tau+u \given Z_\tau = A ] = r \int_\tau^{\tau+u} (1-p(X_s,s)) ds \\
        % \E[ T_A(\tau) &> \tau+u \given Z_\tau = B ] = r \int_\tau^{\tau+u} p(X_s,s) ds .
    \end{aligned}
\end{align}
In the first expression, giving the distribution of how the lineage location changes,
the first term ($\sigma dB_\tau$) is Brownian noise driven by dispersal,
and the second is the mean displacement, which moves the lineage ``downhill'' towards its selected allele's ancestral range on either $-\log p$ or $-\log (1-p)$,
depending on which selected allele the lineage is linked to.
(In two dimensions, $d/dx$ is replaced by the gradient.)
\plr{
The second expression says that the probability a lineage on the $A$ background at $x$
recombines onto the $B$ background
is equal to the recombination rate, $r$, multiplied by the local proportion of $B$ alleles, $(1-p(x,\tau))$,
per generation.
}
% \plr{
% The second expression gives the probability that a lineage on the $A$ background at time $\tau$
% does not switch to the $B$ background across $u$ generations,
% i.e., the probability that there was no recombination between the lineage and the selected site in those $u$ generations.
% Since the rate of recombination between the two is $r$,
% and the probability the lineage is in a heterozygote at time $s$ 
% is equal to the local proportion of $B$ alleles, $1-p(X_s,s)$,
% the mean number of recombinations in heterozygotes across that time period is $r \int_\tau^{\tau+u} (1-p(X_s,s)) ds$;
% assuming recombinations are Poisson, the probability of no such recombinations is the expression given.
% The third expression is the same, except for switching $B$ to $A$.
% }
% gives the rate at which a lineage switches from an $A$ to $B$ background as a result of recombination,
% which occurs at rate $r$ multiplied by the local proportion of $B$ alleles, which is $(1-p)$.
% The third expression is the corresponding rate for switching from $B$ to $A$.
